Schema Evolution in Interactive Programming Systems

### 2024-08-25 paper.tex
DIvergence creates new feedback channels

This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing the hellscape of manual copy \& paste. But meeting this challenge may require a basic change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application itself offers
version control affordances to the user. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers. The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.

>>>>

This challenge envisions a new feedback channel that end users have not experienced before: the ability to frictionlessly exchange variations and elaborations they make to documents, replacing manual copy \& paste. But meeting this challenge may require a  change of perspective in both live programming and local-first software.
Live programming must move from being a solitary activity to a collaborative workflow, and further where the collaboration is not just on editing source code but on live integrated code and data,
as in the Smalltalk/Lisp images of old.
%\footnote{If only we had this in the 90's when Smalltalk had a shot at the mainstream!}
Local-first software must move from automatically converging data replicas
to also interactively managing long-term divergence. That implies either application state
is being directly exposed to the user, or the application itself offers
version control affordances to the user. We encourage both communities to think outside the box of Git,
which has proven to baffle not only end-users but also a substantial fraction of developers. The Divergence challenge may be as much a technical challenge as a conceptual challenge, asking us to reconsider the entrenched boundaries between technologies in the stack.
